---
title: "Getting Started with Stan"
author: "Dan Ovando"
institute: "EcoDataScience Presentation"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rstan)
library(rstanarm)
library(bayesplot)
library(tidybayes)
library(gapminder)

```


# Objectives for Workshop

1. Basics of Bayes
  - What is it and why use it

2. Bayesian regression with `rstanarm`

  - Diagnosing with `shinystan`
  
  - Getting and plotting results with `tidybayes`/`bayesplot`

3. Roll your own
  - Writing your own models in stan
  - Not going to address `brms`
  

---

class: center, middle, inverse
# Basics of Bayes


---

# Basics of Bayes


Bayes Theorem:

$$p(model|data) = \frac{p(data|model) \times p(model)}{p(data)} $$


---

# Basics of Bayes


Include cookies graphic from dissertation


---

# Bayesian vs. Frequentist

This is a deep topic that we dont' have time to get into! But from an ecological perspectice Bayes has some nice features


.pull-left[

### Bayes

What is the probability of a model given the data?

- e.g. conditional on my model and data, how likely is it that a stock is overfished?

Clean way to bring in prior knowledge

But, means you have to think about priors

] .pull-right[

### Frequentist

How likely are the observed data if a given model is true?

What you probably learned in stats classes.

Can't really say anything about the probability of a parameter. 
No need to think about priors


]


---


# Basics of Bayes - MCMC


Big reason that Bayes wasn't historically used as much was that except for some specific cases no analytical solution (and debates about priors)
  
  - Frequentist can usually solve for the answer


Started to change when computers made Markov Chain Monte Carlo (MCMC) practical
  - Monte Carlo - try lots of different numbers
  
  - Markov Chain - an elegent way of choosing those numbers

---

# Basics of Bayes - MCMC


MCMC can be theoretically shown to always converge! 

Sounds like magic, buy basically says, "if you try every number in existence, you'll find solution"

The trick then is finding an efficient way to explore the parameter space. 

XX MCMC psuedocode XX


---


# Enter Stan (and Hamilton!)

Fuill disclosure: I've never seen or heard Hamilton, but I hear it's popular. 

Stan uses a very elegant method called Hamiltonian Monte Carlo with a No-U-turn sampler (NUTs) to help Bayesian models converge quickly with relatively clear diagnostics. 

We don't have time to go into it, but trust me, it's cool. 

XX video of stan fit XX

---


class: center, middle, inverse
# Bayesian regression with `rstanarm`


---

# Bayesian regression with `rstanarm`

Bayesian methods traditionally required writing your own code of the model. Made running say a regression pretty annoying. 

`rstanarm` was developed to reduce the "it's a pain" reason for not using Bayesian regressions

  - Modification of classic "applied regression modeling" (`arm`) package

---


# Bayesian regression with `rstanarm`

.pull-left[
```{r}

bayes_model <- rstanarm::stan_glm("lifeExp ~ gdpPercap",
                                 data = gapminder,
                                 refresh = 0)

summary(bayes_model)
```
] .pull-right[


```{r}


freq_model <- lm("lifeExp ~ gdpPercap",
                   data = gapminder)

summary(freq_model)
```

]



---

# Setting Priors

Determining priors is an art and a science

`rstanarm` has some pretty clever selectors for weakly informative priors, but good practice to set them yourselves. 

https://mc-stan.org/rstanarm/articles/priors.html

https://mc-stan.org/rstanarm/articles/priors.html

View default priors
---

# Setting Priors


Example of an imformative priors

---

# Diagnosing Models

Divergences

Maxtreedepth

Energy

---

# Diagnosing with shinystan


---


# Model Comparison with `loo`


---


# Get things out with `tidybayes`


---

class: center, middle, inverse
# Writing models in `stan`


---

# Writing models in `stan`

`rstanarm` is great and helps make Bayesian regressions just as easy as `lm/glm/gamm` etc. 

Lots of times though, we want to move beyond linear regression. Bayesian estimation through HMC is particularly good at estimating really complex models. 

For that, you may need to write your own stan code
  - [brms](https://paul-buerkner.github.io/brms/) is a great package for writting non-linear models in stan without writing the stan code. Not going to cover here
  
  

---


# Anatomy of a stan model

breakdown of the components of a stan model

---


# General stan syntax

C++ like (but no zero indexing!)

Lots of nice vectorization built in

---

# data block


---


# parameter block


---

# transformed parameter block


---


# model block


---

# generated quantities


---


# Examining your model


---


# Using your model to make predictions


---

# Where to learn more?

Vignettes, bugs examples, etc. 

---




